---
title: Density surface models
author: David L Miller

---

# Load some data and packages

```{r load-data-packages}
library(Distance)
library(dsm)
library(ggplot2)
library(knitr)
load("sperm-data.RData")
load("df-models.RData")
```

# Fit some models


## quasipoisson

warnings about fitting linear terms (and terms that up tick) for prediction/extrapolation.

try models with bivariate location term and separate additive terms

```{r fit-models-qp-xy}
k2 <- 30
k1 <- 10

dsm.big.qp.xy <- dsm(count~s(x, y, k=k2, bs="ts")+
                           s(Depth, k=k1, bs="ts")+
#                           s(DistToCAS, k=k1, bs="ts")+
                           s(SST, k=20, bs="ts")+
#                           s(EKE, k=k1, bs="ts")+
                           s(NPP, k=k1, bs="ts"),
                     family=quasipoisson(),
                     df_hr, effort, obs, method="REML")
summary(dsm.big.qp.xy)
```

checking plots

```{r qp-check-xy, fig.width=6, fig.height=6}
gam.check(dsm.big.qp.xy)
```

plot smooths

```{r qp-plot-xy, fig.width=6, fig.height=6}
plot(dsm.big.qp.xy, pages=1, scale=0)
```

```{r fit-models-qp}
dsm.big.qp <- dsm(count~s(x, k=k1, bs="ts")+
                        s(y, k=k1, bs="ts")+
                        s(Depth, k=k1, bs="ts")+
#                        s(DistToCAS, k=k1, bs="ts")+
                        s(SST, k=20, bs="ts"),#+
#                        s(EKE, k=k1, bs="ts"),#+
#                        s(NPP, k=k1, bs="ts"),
               family=quasipoisson(),
               df_hr, effort, obs, method="REML")
summary(dsm.big.qp)
```


note that NPP goes out first with `x`,`y` additive but stays in for bivariate terms

checking plots

```{r qp-check, fig.width=6, fig.height=6}
gam.check(dsm.big.qp)
```

plot that

```{r qp-plot, fig.width=6, fig.height=6}
plot(dsm.big.qp, pages=1, scale=0)
```


## Tweedie


`DistToCAS` and `NPP` here give good examples of `bs="ts"` working.

```{r fit-models-tw-xy}
dsm.big.tw.xy <- dsm(count~ s(x,y, k=k2, bs="ts")+
                         s(Depth, k=k1, bs="ts"),#+
#                         s(DistToCAS, k=k1, bs="ts")+
#                         s(SST, k=k1, bs="ts")+
#                         s(EKE, k=k1, bs="ts"),#+
#                         s(NPP, k=k1, bs="ts"),
               family=tw(),
               df_hr, effort, obs, method="REML")
summary(dsm.big.tw.xy)
```

```{r tw-check-xy, fig.width=6, fig.height=6}
gam.check(dsm.big.tw.xy)
```

```{r tw-rq-check-xy, fig.width=6, fig.height=6}
rqgam.check(dsm.big.tw.xy)
```

```{r tw-plot-xy, fig.width=6, fig.height=6}
plot(dsm.big.tw.xy, pages=1, scale=0)
```

```{r fit-models-tw}
dsm.big.tw <- dsm(count~ #s(x, k=k1, bs="ts")+
#                         s(y, k=k1, bs="ts")+
                         s(Depth, k=k1, bs="ts")+
#                         s(DistToCAS, k=k1, bs="ts")+
#                         s(SST, k=k1, bs="ts")+
#                         s(EKE, k=k1, bs="ts")+
                         s(NPP, k=k1, bs="ts"),
               family=tw(),
               df_hr, effort, obs, method="REML")
summary(dsm.big.tw)
```

```{r tw-check, fig.width=6, fig.height=6}
gam.check(dsm.big.tw)
```

```{r tw-rq-check, fig.width=6, fig.height=6}
rqgam.check(dsm.big.tw)
```

```{r tw-plot, fig.width=6, fig.height=6}
plot(dsm.big.tw, pages=1, scale=0)
```

## negative binomial

```{r fit-models-nb-xy}
dsm.big.nb.xy <- dsm(count~s(x,y, k=k2, bs="ts")+
                        s(Depth, k=k1, bs="ts"),#+
#                        s(DistToCAS, k=k1, bs="ts")+
#                        s(SST, k=k1, bs="ts")+
#                        s(EKE, k=k1, bs="ts"),#+
#                        s(NPP, k=k1, bs="ts"),
               family=nb(),
               df_hr, effort, obs, method="REML")
summary(dsm.big.nb.xy)
```

```{r nb-check-xy, fig.width=6, fig.height=6}
gam.check(dsm.big.nb.xy)
```

```{r nb-rq-check-xy, fig.width=6, fig.height=6}
rqgam.check(dsm.big.nb.xy)
```

```{r nb-plot-xy, fig.width=6, fig.height=6}
plot(dsm.big.nb.xy, pages=1, scale=0)
```




```{r fit-models-nb}
dsm.big.nb <- dsm(count~#s(x, k=k2, bs="ts")+
                        s(y, k=k2, bs="ts")+
                        s(Depth, k=k1, bs="ts"),#+
#                        s(DistToCAS, k=k1, bs="ts")+
#                        s(SST, k=k1, bs="ts")+
#                        s(EKE, k=k1, bs="ts"),#+
#                        s(NPP, k=k1, bs="ts"),
               family=nb(),
               df_hr, effort, obs, method="REML")
summary(dsm.big.nb)
```

```{r nb-check, fig.width=6, fig.height=6}
gam.check(dsm.big.nb)
```

```{r nb-rq-check, fig.width=6, fig.height=6}
rqgam.check(dsm.big.nb)
```

```{r nb-plot, fig.width=6, fig.height=6}
plot(dsm.big.nb, pages=1, scale=0)
```



## Model summary


```{r summarize-models}
summarize_dsm <- function(model){

  summ <- summary(model)

  data.frame(response = model$family$family,
             terms    = paste(rownames(summ$s.table), collapse=", "),
             AIC      = AIC(model),
             REML     = model$gcv.ubre,
             Deviance = round(summ$dev.expl*100,2)
            )

}
library(plyr)
summary_table <- ldply(list(dsm.big.qp,
                             dsm.big.qp.xy,
                             dsm.big.tw,
                             dsm.big.tw.xy,
                             dsm.big.nb,
                             dsm.big.nb.xy),
                        summarize_dsm)
summary_table <- as.data.frame(summary_table)
```

```{r print-table, results="asis"}
summary_table <- summary_table[order(summary_table$REML, decreasing=TRUE),]
kable(summary_table)
```

## Final models

```{r plot-qq}
par(mfrow=c(2,3))
qq.gam.p <- function(x){qq.gam(x, main=paste0(substitute(x)))}
qq.gam.p(dsm.big.qp)
qq.gam.p(dsm.big.qp.xy)
qq.gam.p(dsm.big.tw)
qq.gam.p(dsm.big.tw.xy)
qq.gam.p(dsm.big.nb)
qq.gam.p(dsm.big.nb.xy)
```

Looking at the results we see that 


Comparing the depth terms we see a similar story:

```{r depth-plots, fig.width=5, fig.height=8}
par(mfrow=c(3,2))
plot_depth <- function(model){
  plot(model, select=which(grepl("Depth", unlist(strsplit(as.character(terms(model$formula)[[3]]), "\\+"))))-1, scale=0, main=paste0(substitute(model)))
  }
plot_depth(dsm.big.qp)
plot_depth(dsm.big.qp.xy)
plot_depth(dsm.big.tw)
plot_depth(dsm.big.tw.xy)
plot_depth(dsm.big.nb)
plot_depth(dsm.big.nb.xy)
```

```{r save-models}
save(dsm.big.qp,
     dsm.big.qp.xy,
     dsm.big.tw,
     dsm.big.tw.xy,
     dsm.big.nb,
     dsm.big.nb.xy,
     file="count-models.RData")
```
