---
title: Estimating variance
author: David L Miller

---

Now we've fitted some models and estimated abundance, we can estimate the variance associated with the abundance estimate (and plot it).

# Aims

By the end of this practical, you should feel comfortable:


# Load packages and data

```{r load-packages}
library(dsm)
library(raster)
library(ggplot2)
library(viridis)
library(plyr)
library(knitr)
```

Load the models and prediction grid:
```{r load-models}
load("dsms.RData")
load("dsms-xy.RData")
load("predgrid.RData")
```

# Estimation of variance

Depending on the model response (count or Horvitz-Thompson) we can use either `dsm.var.prop` or `dsm.var.gam`, respectively. `dsm_nb_xy_ms` doesn't include any covariates at the observer level in the detection function, so we can use the variance propagation method and estimate the uncertainty in detection function parameters in one step.

```{r varest}
# need to remove the NAs as we did when plotting
predgrid_var <- predgrid[!is.na(predgrid$Depth),]
# now estimate variance
var_nb_xy_ms <- dsm.var.prop(dsm_nb_xy_ms, predgrid_var, off.set=predgrid_var$off.set)
```
To summarise the results of this variance estimate:
```{r varest-summary}
summary(var_nb_xy_ms)
```

Try this out for some of the other models you've saved.

# Summarise multiple models

We can again summarise all the models, as we did with the DSMs and detection functions, now including the variance:

```{r summarise-var}
summarize_dsm_var <- function(model, predgrid){

  summ <- summary(model)
  
  vp <- summary(dsm.var.prop(model, predgrid, off.set=predgrid$off.set))
  unconditional.cv.square <- vp$cv^2
  asymp.ci.c.term <- exp(1.96*sqrt(log(1+unconditional.cv.square)))
  asymp.tot <- c(vp$pred.est / asymp.ci.c.term,
                 vp$pred.est,
                 vp$pred.est * asymp.ci.c.term)

  data.frame(response = model$family$family,
             terms    = paste(rownames(summ$s.table), collapse=", "),
             AIC      = AIC(model),
             REML     = model$gcv.ubre,
             "Deviance_explained" = paste0(round(summ$dev.expl*100,2),"%"),
             "lower_CI" = round(asymp.tot[1],2),
             "Nhat" = round(asymp.tot[2],2),
             "upper_CI" = round(asymp.tot[3],2)
             )
}
```
```{r var-table}
# make a list of models (add more here!)
model_list <- list(dsm_nb_xy, dsm_nb_x_y, dsm_nb_xy_ms, dsm_nb_x_y_ms)
# give the list names for the models, so we can identify them later
names(model_list) <- c("dsm_nb_xy", "dsm_nb_x_y", "dsm_nb_xy_ms", "dsm_nb_x_y_ms")
per_model_var <- ldply(model_list, summarize_dsm_var, predgrid=predgrid_var)
```

```{r print-table}
kable(per_model_var)
```


# Plotting

We can plot a map of the coefficient of variation, but we first need to estimate the variance per prediction cell, rather than over the whole area. This calculation takes a while!

```{r per-cell-var}
# use the split function to make each row of the predictiond data.frame into
# an element of a list
predgrid_var_split <- split(predgrid_var, 1:nrow(predgrid_var))
var_split_nb_xy_ms <- dsm.var.prop(dsm_nb_xy_ms, predgrid_var_split, off.set=predgrid_var$off.set)
```
Now we have the per-cell coefficients of variation, we assign that to a column of the prediction grid data and plot it as usual:

```{r varest-map-obs}
predgrid_var_map <- predgrid_var
cv <- sqrt(var_split_nb_xy_ms$pred.var)/unlist(var_split_nb_xy_ms$pred)
predgrid_var_map$CV <- cv
p <- ggplot(predgrid_var_map) +
       geom_tile(aes(x=x, y=y, fill=CV, width=10*1000, height=10*1000)) +
       scale_fill_viridis() +
       coord_equal() +
       geom_point(aes(x=x,y=y, size=count), data=dsm_nb_xy_ms$data[dsm_nb_xy_ms$data$count>0,])
print(p)
```

Note that here we overplot the segments where sperm whales were observed (and scale the size of the point according to the number observed), using `geom_point()`.

Actually this is more useful when we overplot the effort, to better see where we sampled:

```{r varest-map-effort}
#p <- ggplot(predgrid_var_map) +
#       geom_tile(aes(x=x, y=y, fill=CV, width=10*1000, height=10*1000)) +
#       scale_fill_viridis() +
#       coord_equal() +
#       geom_path(aes(x=x, y=y,))
#print(p)
```


Try this with the other models you fitted and see what the differences are between the maps of coefficient of variation.


# Save the uncertainty maps to raster files

As with the predictions, we'd like to save our uncertainty estimates to a raster layer so we can plot them in ArcGIS. Again, this involves a bit of messing about with the data format before we can save.

```{r savecv-raster}
# setup the storage for the cvs
cv_raster <- raster(nrow=nrow(predictorStack), ncol=ncol(predictorStack),
                    ext = predictorStack@extent, crs=predictorStack@crs)
# we removed the NA values to make the predictions and the raster needs them
# so make a vector of NAs, and insert the CV values...
cv_na <- rep(NA, nrow(predgrid))
cv_na[!is.na(predgrid$Depth)] <- cv
# put the values in, making sure they are numeric first
cv_raster <- setValues(cv_raster, cv_na)
# add that to the stack
save_raster <- stack(c(predictorStack, cv_raster))
# name the new, last, layer in the stack
names(save_raster)[length(names(save_raster))] <- "CV_nb_xy"
```
We can then save that object to disk as a raster file:
```{r write-raster}
writeRaster(save_raster, "cv_raster.img", datatype="FLT4S")
```




